---
title: "Analysis of Global Reading Proficiency and Life Expectancy"
subtitle: "presents an analysis of the reading proficiency across various countries using data from UNICEF indicators."
author: "Bijoy Krishna Kalita"
date: "2024-04-25"
format:
  html:
    theme: sketchy
    toc: true
    
execute: 
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}

#libraries
library(tidyverse)
library(ggplot2)
library(maps)
library(dplyr)
library(countrycode)
library(plotly)

#data
indicator1 <- read.csv('unicef_indicator_1.csv')
indicator2 <- read.csv('unicef_indicator_2.csv')
metadata <- read.csv('unicef_metadata.csv')

#transformations
indicator <- bind_rows(indicator1, indicator2)
indicator_2018 <- indicator1 %>%
  filter(time_period == 2018)
```

## Global Reading Proficiency Map
**Figure 1:** 


> **Overview:** This map provides an intuitive visual representation of reading proficiency across the globe for 2018. Each country's color intensity—from pale to deep orange—reflects its literacy rates, offering a quick visual assessment of educational strengths and needs.

#### Key Insights:
- **Global Leaders:** Countries in dark orange have achieved exemplary educational outcomes.
- **Areas for Improvement:** Light orange shades pinpoint regions needing urgent educational reforms and support.

This visualization is essential for identifying disparities and targeting interventions effectively, shaping policies that focus on underserved areas.


```{r}
world_map <- map_data("world")
indicator_map <- full_join(world_map, indicator_2018, by = c("region" = "country"), relationship = "many-to-many")
# World Map Chart with more detailed aesthetics
ggplot(data = indicator_map, aes(x = long, y = lat, group = group, fill = obs_value)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "#ffb671", high = "#f88501", na.value = "#ffd7b7") +
  
  #scale_fill_viridis_c(option = "C", direction = -1, name = "Reading Proficiency (%)") +
  labs(
    title = "Global Reading Proficiency at Lower Secondary Level",
    subtitle = "percentage of reading proficiency for the year 2018 is shown",
    x = "",
    y = "",
    ) +
  theme_classic() +
  theme(legend.position = "bottom")
```


## Top 10 Countries in Reading Proficiency
**Figure 2:** 

**Detailed Analysis:**
Each bar represents a country, uniquely colored to highlight its position among global leaders in literacy. This chart does not just celebrate high achievers but also serves as a motivation for other nations to enhance their educational strategies.

#### Top Performers:
- **Countries Exceeding 80% Proficiency:** These are benchmark nations where a significant majority of students reach basic reading competencies by the end of lower secondary school.

The selection of vibrant colors enhances visual appeal and draws attention to each country's educational excellence.


```{r}
top_countries <- indicator1 %>%
  filter(sex == "Total", obs_value > 80 & obs_value < 99) %>% 
  arrange(desc(obs_value)) %>%
  slice_head(n = 10)



custom_colors <- c("#974700", "#ff7f0e", "#f88501", "#f7a156", "#e49a59", "#f0b27d", "#ffa775", "#f8bc99", "#f8bc99", "#f8bc99" ) #defining custom color sets first
ggplot(top_countries, aes(x = reorder(country, obs_value), y = obs_value, fill = country)) +
  geom_col(show.legend = FALSE, width = 0.5) +
 # coord_flip() +
  scale_y_continuous(limits = c(82, 90), breaks = seq(82, 90, by = 1),  oob = scales::oob_squish) +
  
 
  labs(title = "Showing Top 10 Countries by Reading Proficiency", x = "Countries", y = "Proportion (%)",
       subtitle = "These are the nations where more than 80% of students graduate from lower secondary school with at least a basic reading competency. ") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  scale_fill_manual(values = custom_colors)
```

## Gender Disparities in Reading Proficiency
**Figure 3:**

> **Context:** This scatter plot breaks down reading proficiency by gender, with orange for males and teal for females, across various countries. Linear models reveal trends and disparities.

#### Observations:
- **Visible Trends:** Highlighting gender-specific educational achievement and gaps.
- **Policy Implications:** The data prompts deeper investigations into socio-cultural and economic factors influencing these disparities, guiding targeted educational policies.

This visualization is a crucial tool for understanding and addressing gender inequalities in global education.



```{r}
scatter_data <- indicator1 %>%
  group_by(country, sex) %>%
  filter(any(obs_value > 50)) %>%
  slice_head(n = 25)

ggplot(scatter_data, aes(x = country, y = obs_value, color = sex)) +
  geom_point(alpha = 50, size = 3.8, color = "orange") +
  #geom_smooth(method = "lm") +
  #geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", size = 2) +
  
  geom_smooth(method = "lm", aes(group = sex), se = FALSE, 
              color = "darkorange", size = 1, linetype = "solid") +
  
  labs(title = "Reading Proficiency",
       subtitle = "showing something...............",
       x = "Countries", y = "Reading Proficiency (%)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "",
        panel.grid.major = element_blank(),   
    panel.grid.minor = element_blank(),  
    panel.background = element_blank())
```


## Life Expectancy Trends (2010-2019)
**Figure 4:** 


**Examination of Health Outcomes:**
This line graph provides a detailed analysis of life expectancy over a decade in selected developed countries. Each line's unique color aids in tracking the health progress or decline of a specific country.

#### Health Insights:
- **Correlation with Education:** Insights into how educational levels potentially influence health outcomes.
- **Yearly Trends:** Detailed observations of each country's health improvements or setbacks over the years.

This graph is not only about health metrics but also about understanding the broader narrative of societal well-being tied to education.




```{r}
country_list <- c("Australia", "Canada", "France", "Germany", "Italy", 
                         "Japan", "New Zealand", "Norway", "United Kingdom", "United States")
country_colors <- c("Australia" = "#E41A1C", "Canada" = "#377EB8", "France" = "#4DAF4A", 
                    "Germany" = "#984EA3", "Italy" = "#FF7F00", "Japan" = "#FFFF33",
                    "New Zealand" = "#A65628", "Norway" = "#F781BF", "United Kingdom" = "#999999",
                    "United States" = "#66C2A5")


filtered_metadata <- metadata %>%
  filter(country %in% country_list, time_period >= 2010 & time_period <= 2020) %>%
  select(country, time_period, Life_expectancy_at_birth_total_years)

filtered_metadata$time_period <- as.integer(filtered_metadata$time_period)



ggplot(filtered_metadata, aes(x = time_period, y = Life_expectancy_at_birth_total_years, group = country, color = country)) +
  geom_line(size = 0.8) +  
  geom_point() + 
  scale_color_manual(values = country_colors) +
  scale_x_continuous(breaks = seq(min(filtered_metadata$time_period), max(filtered_metadata$time_period), 1)) +
  labs(
    title = "Life Expectancy Trends from 2010 to 2019",
    subtitle = "Comparison among Selected Developed Countries",
    x = "Year",
    y = "Life Expectancy at Birth (Years)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(face = "italic", size = 12),
    axis.text.x = element_text(angle = 0, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 12)
  )
```

## Conclusion & Recommendation
Together, these visualizations form a narrative that links educational achievement with broader socio-economic outcomes, such as health and quality of life. By comparing countries and highlighting disparities in education and health, your story emphasizes the need for a focused approach to improve educational systems and healthcare policies globally, suggesting that investment in education could be directly beneficial to societal health and longevity.

These insights derived from your data not only inform policymakers and educators but also engage the general public in understanding the critical challenges and achievements in global education and health.



